---
title: "STA 160 Project"
author: "Gabriel Jones"
date: "2024-11-04"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Libraries**

```{r echo=F, results=F}
list.of.packages <- c("MASS", "dplyr", "tidyverse", "ggplot2", "leaps", "boot", 
                      "knitr", "hrbrthemes", "RColorBrewer", "paletteer", 
                      "glmnet", "corrplot")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
lapply(list.of.packages, library, character.only = TRUE)
```
**Load Data**

```{r}
data <- read.csv("Life Expectancy Data.csv")
data <- na.omit(data)
```

```{r}
glimpse(data)
```

```{r}
count(data,Country)
```

**Yearly data points**

```{r}
data%>%
  ggplot(aes(Year))+
  geom_bar()
```

**Overall Life Expectancy by Year**

```{r}
data%>%
  mutate(Year=as.factor(Year))%>%
  ggplot(aes(Year, Life.expectancy))+
  geom_boxplot()
```

**Correlation Matrix**

```{r}
dataNumeric <- data%>%
  select(-"Country", -"Year", -"Status")%>%
  na.omit()

corrplot(cor(dataNumeric), method="color")
```

High Correlations include:

-   Life expectancy & Adult Mortality (-0.70)

-   Life expectancy & income (.72)

-   Life expectancy & schooling (.72)

-   GDP & percentage expenditure (.95)

-   Income & schooling (.78) - thinness vars (.92)

-   Diphtheria & Polio (0.61)

-   Under five deaths & population (.65)

-   infant deaths & population (0.67)

-   Under five deaths & infant deaths (0.99)

Vars of interest may include: *Adult Mortality*, *HIV AIDS*, *Income*, and *Schooling.*

If the above are used, consider dropping either income or schooling

**Life Expectancy by Country Summary Statistics**

```{r}
# Numerical
sumstat <- data.frame("Country"=c(),
                      "Min."=c(),
                      "X1st.Qu."=c(),
                      "Median"=c(),
                      "Mean"=c(),
                      "X3rd.Qu."=c(),
                      "Max."=c())
for (c in unique(data$Country)){
  country.dat <- data$Life.expectancy[data$Country==c]
  r <- cbind(data.frame("Country"=c(c)), data.frame(as.list(summary(country.dat))))
  sumstat <- rbind(sumstat, r)
}
kable(sumstat)

# Graphical
data%>%
  ggplot(aes(x=Country, y=Life.expectancy))+
  geom_boxplot()+
  scale_x_discrete(guide = guide_axis(angle = 70))+
  theme(axis.text.x=element_text(size=7))
```

**Find Life Expectancy Outliers**

```{r}
out.dat <- data[FALSE,]

for (c in unique(data$Country)){
  country.dat <- data$Life.expectancy[data$Country==c]
  lowerq = quantile(data$Life.expectancy[data$Country==c])[2]
  upperq = quantile(data$Life.expectancy[data$Country==c])[4]
  iqr <- IQR(data$Life.expectancy[data$Country==c])
  upper.out = (iqr * 1.5) + upperq
  lower.out = lowerq - (iqr * 1.5)
  for (v in country.dat){
    if (v <= lower.out | v >= upper.out){
      out.dat <- rbind(out.dat, filter(data, Country==c, Life.expectancy==v))
    }
  }
}
kable(head(out.dat, 10))
```

**QQ plots**

```{r}
for (i in 1:length(colnames(dataNumeric))){
  qqnorm(dataNumeric[,i], main = paste0("Normal Q-Q Plot ", colnames(dataNumeric)[i]))
}
```

-   *Adult Mortality* - Slightly Right Skewed

-   *HIV Aids* - Very Right Skewed, many zeros

-   *Income* - Slightly Left Skewed, many Zeros

-   *Schooling* - Normal

**Scatterplots**

```{r}
for (i in 2:length(colnames(dataNumeric))){
  plot(dataNumeric[,i], dataNumeric[,1], 
       main = paste0(colnames(dataNumeric)[i], "Scatterplot"),
       xlab = colnames(dataNumeric)[i],
       ylab = colnames(dataNumeric)[1])
}
```

Trying to figure out cause of Adult Mortality Grouping

```{r}
data%>%
  ggplot(aes(Adult.Mortality, Life.expectancy, color=Status))+
  geom_point()
```

**Residual Plots**

```{r}
for (i in 2:length(colnames(dataNumeric))){
  model <- lm(dataNumeric[,1] ~ dataNumeric[,i])
  res <- resid(model)
  plot(fitted(model), res,
       main = paste0(colnames(dataNumeric)[i], " Residual Plot"),
       xlab = "Residuals",
       ylab = "Fitted Values")
  abline(0,0) 
}
```

**Life Expectancy grouped by Status**

```{r}
data%>%
  ggplot(aes(Status, Life.expectancy))+
  geom_boxplot()

data%>%
  mutate(Year=as.factor(Year))%>%
  ggplot(aes(Year, Life.expectancy))+
  geom_boxplot()+
  facet_wrap(vars(Status))+
  scale_x_discrete(guide = guide_axis(angle = 45))
```

**Adult Mortality grouped by Status**

```{r}
data%>%
  ggplot(aes(Status, Adult.Mortality))+
  geom_boxplot()

data%>%
  mutate(Year=as.factor(Year))%>%
  ggplot(aes(Year, Adult.Mortality))+
  geom_boxplot()+
  facet_wrap(vars(Status))+
  scale_x_discrete(guide = guide_axis(angle = 45))
```

**HIV/AIDS grouped by Status**

```{r}
data%>%
  ggplot(aes(Status, HIV.AIDS))+
  geom_boxplot()

data%>%
  mutate(Year=as.factor(Year))%>%
  ggplot(aes(Year, HIV.AIDS))+
  geom_boxplot()+
  facet_wrap(vars(Status))+
  scale_x_discrete(guide = guide_axis(angle = 45))
```

**Schooling grouped by Status**

```{r}
data%>%
  ggplot(aes(Status, Schooling))+
  geom_boxplot()

data%>%
  mutate(Year=as.factor(Year))%>%
  ggplot(aes(Year, Schooling))+
  geom_boxplot()+
  facet_wrap(vars(Status))+
  scale_x_discrete(guide = guide_axis(angle = 45))
```
