---
title: "STA 160 Project"
author: "Gabriel Jones"
date: "2024-11-04"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Libraries**
```{r echo=F}
list.of.packages <- c("MASS", "dplyr", "tidyverse", "ggplot2", "leaps", "boot", 
                      "knitr", "hrbrthemes", "RColorBrewer", "paletteer", 
                      "glmnet", "corrplot", "ggpairs")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
lapply(list.of.packages, library, character.only = TRUE)
```


```{r}
data <- read.csv("Life Expectancy Data.csv")
data <- na.omit(data)
```

```{r}
count(data,Country)
```

```{r}
data%>%
  ggplot(aes(Year))+
  geom_bar()
```

Correlation Matrix
```{r}
dataNumeric <- data%>%
  select(-"Country", -"Year", -"Status")%>%
  na.omit()

corrplot(cor(dataNumeric), method="color")
```
High Correlations include:
- Life expectancy & Adult Mortality (-0.70)
- Life expectancy & income (.72)
- Life expectancy & schooling (.72)
- GDP & percentage expenditure (.95)
- Income & schooling (.78)
- thinness vars (.92)
- Diphtheria & Polio (0.61)
- Under five deaths & population (.65)
- infant deaths & population (0.67)
- Under five deaths & infant deaths (0.99)

Life Expectancy by Country Boxplot
```{r}
data%>%
  na.omit()%>%
 # group_by(Country)%>%
  ggplot(aes(x=Country, y=Life.expectancy))+
  geom_boxplot()+
  scale_x_discrete(guide = guide_axis(angle = 70))
```
Find Outliers

```{r}
out.dat <- data[FALSE,]

for (c in unique(data$Country)){
  country.dat <- data$Life.expectancy[data$Country==c]
  lowerq = quantile(data$Life.expectancy[data$Country==c])[2]
  upperq = quantile(data$Life.expectancy[data$Country==c])[4]
  iqr <- IQR(data$Life.expectancy[data$Country==c])
  upper.out = (iqr * 1.5) + upperq
  lower.out = lowerq - (iqr * 1.5)
  for (v in country.dat){
    if (v <= lower.out | v >= upper.out){
      out.dat <- rbind(out.dat, filter(data, Country==c, Life.expectancy==v))
    }
  }
}
out.dat
```

QQ plots
```{r}
for (i in 1:length(colnames(dataNumeric))){
  qqnorm(dataNumeric[,i], main = paste0("Normal Q-Q Plot ", colnames(dataNumeric)[i]))
}
```

Scatterplots
```{r}
for (i in 2:length(colnames(dataNumeric))){
  plot(dataNumeric[,i], dataNumeric[,1], 
       main = paste0(colnames(dataNumeric)[i], "Scatterplot"),
       xlab = colnames(dataNumeric)[i],
       ylab = colnames(dataNumeric)[1])
}
```

Residual Plots
```{r}
for (i in 2:length(colnames(dataNumeric))){
  model <- lm(dataNumeric[,1] ~ dataNumeric[,i])
  res <- resid(model)
  plot(fitted(model), res,
       main = paste0(colnames(dataNumeric)[i], " Residual Plot"),
       xlab = "Residuals",
       ylab = "Fitted Values")
  abline(0,0) 
}
```

